---
title: "Data frame manipulation with tidyr"
author: "Espen Rosenquist"
date: "6 juni 2017"
output: html_document
---
```{r setup, include=FALSE}
library("knitr")
knitr::opts_chunk$set(echo = TRUE)
```

## Prerequisites
We need some sample data, which we save in our `Data` folder. 

```{r}
packagelist <- c("readxl", "tidyverse")
lapply(packagelist, library, character.only = TRUE)
```


```{r, error=TRUE, eval = FALSE}
# See the etherpad - but really doesn't work on Windows. Must download manually or find other package/parameters to download with
download.file(url = "http://docs.google.com/spreadsheet/pub?key=phAwcNAVuyj0NpF2PTov2Cw&output=xlsx", destfile = "Data/indicator gapminder infant_mortality.xlsx")

download.file(url = "http://docs.google.com/spreadsheet/pub?key=phAwcNAVuyj0NpF2PTov2Cw&output=xlsx", destfile = "Data/indicator undata total_fertility.xlsx", method = "wininet")
```

```{r}
raw_fert <- read_excel(path = "./Data/indicator undata total_fertility.xlsx", sheet = "Data")
raw_infantMort <- read_excel(path = "./Data/indicator gapminder infant_mortality.xlsx", sheet = "Data" )
GapMinder <- read_csv("./Data/gapminder-FiveYearData.csv")
```

## Background
There is a frequent need to tidy data from sources. 

The, arguably, best way of organising data is where:

- each variable has its own column
- each observation has its own row
- contained within one dataset

There are two kinds of data:

- **Wide**: each row is an observation and multiple variables containing the same type of data. Human readable
- **Long**: one column for the observed variable and the other columns are ID columns. Computer readable

### challenge
*Is gapminder a purely wide, long, or tidy?*

```{r}
knitr::kable(head(GapMinder))
```
It has three observation values and three ID-columns. It is therefore not wide.
Neither is it long, because it contains different unit measurements in the columns pop, lifeExp, dpPercap.
It is considered tidy.

The two datasets we downloaded are... messy. 
```{r}
knitr::kable(head(raw_fert))
```


**Observe**:

- each years are headers, columns (as integer) -> *it is wide*

We should tidy the data by collect all years as values in a column named `year` and each existing value in a valuecolumn.

## gather

- data set
- columns
- values

```{r}
fert <- raw_fert %>% 
    rename(country=`Total fertility rate`) %>% 
    gather(key = year, value = fert, -country) %>% 
    mutate(year = as.integer(year))
```

