---
title: "tidyverse day 2"
author: "Espen Rosenquist"
date: "7 juni 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, echo = FALSE, message = FALSE}
packagelist <- c("tidyverse", "maps")
lapply(packagelist, library, character.only = TRUE)
```

# Reproducing a graph

## Gapminder plus
```{r}
download.file(url = "https://raw.githubusercontent.com/dmi3kno/SWC-tidyverse/master/data/gapminder_plus.csv", 
              destfile = "Data/gapminder_plus.csv")

gapminder_plus <- read_csv(file = "./Data/gapminder_plus.csv")
```

## Final exam
`*Key parameters for selected Aftrican countries with over 2 million baby deaths in 2007*

- fertility
- gdp per billion
- gdp per capita
- infant mortality
- life expectancy
- population minimum


```{r}
gapminder_plus %>% 
    filter(continent == "Africa") %>% 
    mutate(babiesDead = infantMort/1000*pop) %>% 
    filter(babiesDead>2e6, year == 2007) %>% 
    select(country) %>% 
    left_join(gapminder_plus, by = "country") %>% 
    mutate(babiesDead=infantMort/1000*pop,
           gdb_bln = gdpPercap*pop/1e9,
           pop_mln = pop/1e6) %>% 
    select(-c(continent, pop, babiesDead)) %>%
    gather(key = variables, value = value, -c(country, year)) %>% 
    ggplot() +
    # dot is placeholder for first parameter input - not necessary though
    geom_text(data = . %>% 
                  filter(year == 2007) %>% 
                  group_by(variables) %>% 
                  mutate(max_value = max(value)) %>% 
                  filter(value == max_value), 
              mapping = aes(x = year, 
                            y = value, 
                            label = country, 
                            color = country,
                            size = 10),
              hjust = "right",
              vjust = 0.7) +
    geom_line(aes(x = year, y = value, group = country, color = country)) +
        labs(y = NULL,
         x = "Year",
         title = "Key parameters for selected African countries",
         subtitle = "with over 2 million baby deaths in 2007",
         caption = "Fertility and infant mortality in number of babies per 1000 population, GDP in USD million, GDP per capita in USD, Life expectancy in years, population in million")+
    theme_bw() +
    theme(legend.position = "none") +
    facet_wrap(~variables, scales = "free_y")

```

## Vectors
A vector can be a single number:
```{r}
(x <- 5*6)
```
You can check the type:
```{r}
is.vector(x)
```
You can check it's length:
```{r}
length(x)
```

### Subsetting
You can fill a vector AND read it out by subsetting it. You do this by single brackets.
```{r}
x[2] <- 31
x[2]
x[5] <- 42
```

**Position 0 is not valid**
```{r}
x[0]
```

R is build around vectors, so you can do vector math very easily:
```{r}
x <- 1:10
x^2
```
the non-vector way:
```{r}
z <- vector(mode = mode(x), length = length(x))
for(i in seq_along(x)){
    z[i] <- x[i]^2
}
```

### Recycling
```{r}
x <- 1:5
y <- 3:7
x + y
# z is y, except for position 5:
z <- y[-5]

# You will get a warning message:
x + z

# if one vector in this is a multiple of the other, you will not get a warning
z <- 1:10
x + z
```
Without recycling you would not able to do this:
```{r}
z^x
```

Vectors can be of all types, but only one at a time.

```{r}
str(c("Hello", "workshop", "participants"))
str(c(1:4, 10000, x))
```
R will try to coerce all values to one type.
```{r}
str(c("something", pi, 2:4, pi>3))
str(c(pi, 2:4, pi>3))
str(c(2L:4L, pi>3))
```
But still, you were able to put all kinds of stuff into the vector. So, **be aware**. 
It first resolved every element <- the coercion happens at the end.


```{r}
set.seed(42)
# 10 random floating numbers
w <- rnorm(10)
# indexing is still integer
seq_along(w)
# index of only the negative numbers
which(w < 0)
# the actual negative numbers
w[which(w<0)]
# a logical vector
w < 0

# drop a couple of positions
w[-c(2,5)]
```
## lists
```{r}
x <- list("something", pi, 2:4, pi > 3)
str(x)
```
This list contains:
 - character
 - number
 - numberic vector
 - boolean
 
 A list allows us to name each element:
 
```{r}
x <- list(vegetable = "cabbage",
          number = pi,
          series = 2:4,
          telling = pi > 3)
str(x)
```
Lists can be subset too:
```{r}
# as list
str(x[1])
# as value, by name
str(x$vegetable)
# as value, by index
str(x[[1]])
```
Nested lists:
```{r}
x <- list(vegetable = list("cabbage","spinach", "carrot"),
          number = list(c(pi,0, 2.14, NA)),
          series = list(list(2:4,3:5)),
          telling = pi > 3)
str(x)
```

### Excercise
Extract an element from a complex list.
```{r}
# make a linear model (y ~ independent variable, data):
(mod <- lm(lifeExp ~ gdpPercap, data = gapminder_plus))
# we get a complex list!
str(mod)
```



Subsetting becomes a bitch:
```{r}
# Simple example
# df.residual is a vector
mod$df.residual
mod[["df.residual"]]
mod[[8]]
#qr.qr.first element
mod[["qr"]][["qr"]][1]
mod$qr$qr[1]
mod$qr$qr[1,1]
```
## Rich data
How do we get to rich data without collapsing it?

### Poor data
Summary life exptancy by continent:
```{r}
gapminder_plus %>% 
    group_by(continent) %>% 
    summarise(meanLifeExp = mean(lifeExp),
              minLifeExp = min(lifeExp),
              maxLifeExp = max(lifeExp)) 
```
```{r}
gapminder_plus %>% 
    ggplot() +
    geom_line(mapping = aes(x = year, y = lifeExp, color = continent, group = country))
```

What happened to those outliers? They are outside the trends (hypothesis).
```{r}
gapminder_plus %>% 
    ggplot(aes(x = year, y = lifeExp)) +
    geom_line(aes(color = continent, group = country)) +
    geom_smooth(aes(x = year, y = lifeExp), method = "lm", color = "black") +
    facet_wrap(~continent)
```

Now we want to inspect the residuals, fit linear models for each country.
*Residual: The difference between expected model and observed.*

```{r}
by_country <- gapminder_plus %>% 
    group_by(continent, country) %>% 
    nest()
# not run
# str(by_country$data)
# First data frame:
str(by_country$data[[1]])
```
The third element is a list which is nested.
The first element of the list is a data frame, giving data about the first element in the `by_country`.


### The map function
```{r}
library(purrr)
```

Map is a less ambigous `apply`. It expects a list.

```{r}
map(1:3, sqrt)
map_dbl(1:3, sqrt)
```

Let's make a new complex object (we could just pipe further, but):
```{r}
model_by_country <- by_country %>% 
    mutate(model = purrr::map(data, ~lm(lifeExp ~ year, data = .x)))
```
The ~ specifies a function, while the .x defines where the data is to land. Let's go further and make a summary:
*note: map from purrr does not override the maps package, so we either have to detach the maps or qualify the map function.*
```{r}
model_by_country <- by_country %>% 
    mutate(model = purrr::map(data, ~lm(lifeExp ~ year, data = .x))) %>% 
    mutate(summr = purrr::map(model, broom::glance)) %>% 
    # take out the model data as separate columns
    unnest(summr) %>% 
    # r square = 1 == perfect model -> 0.94 is quite good, so let's sort the data by r.squared
    arrange(r.squared)
head(model_by_country)

model_by_country %>%
    ggplot() +
    geom_jitter(aes(x = continent, y = r.squared, color = continent, group = country)) +
    theme(legend.position = "none")
```
This is getting closer. Let's filter the data and join the interesting countries back into gapminder_plus:
```{r}
low_countries <- model_by_country %>% 
    filter(r.squared < 0.5) %>% 
    select(country)
```

```{r}
world <- map_data("world")  %>% 
    rename(country = region) %>% 
    left_join(gapminder_plus, by = "country")
```

### Assignment
life expectancy per gdpcapita
gapminder_p
```{r}
gapminder_plus %>% 
    ggplot() +
    geom_point(aes(x = log(gdpPercap), y = lifeExp, color = continent, group = country)) +
    theme(legend.position = "none")
```
Life expectancy per gross domestic product per capita
```{r}
by_country <- gapminder_plus %>% 
    group_by(continent, country) %>% 
    nest() %>% 
    mutate(model = purrr::map(data, ~lm(lifeExp ~ log(gdpPercap), data = .x))) %>% 
    mutate(summr = purrr::map(model, broom::glance)) %>% 
    # take out the model data as separate columns
    unnest(summr) %>% 
    # r square = 1 == perfect model -> 0.94 is quite good, so let's sort the data by r.squared
    arrange(r.squared)

by_country %>% 
    filter(r.squared <0.1) %>% 
    select(country, r.squared) %>% 
    left_join(gapminder_plus) %>% 
    ggplot() +
    geom_point(aes(x = log(gdpPercap), y = lifeExp, color = country, group = country, size = r.squared))

```

## Save objects
The (second) last thing to learn today is how to save an object, so that we can pick it up laters. THis is different from CSV or other formats.
It is only readable in R.

```{r}
saveRDS(by_country, "./Data/by_country_tibble.rds")
```

We save the data as CSV:
```{r}
write_csv(gapminder_plus, "./Data/gapminder_plus_for_professor.csv")
```

The very last thing we do is to commit our changes to GitHub.
